---
title: "MiCoFormer V2 Hyperparameter Search Results"
author: "Qingyun Cai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.retina = 2, dpi = 150
)

library(tidyverse)
library(patchwork)
library(yaml)
library(scales)
library(kableExtra)

# -- Path configuration --
ANALYSIS_DIR <- "/Users/kiancai/STA24/CWD/MCFProjet/MiCoFormer/protocols/hparam_search_v2/analysis"
RUN_DIR <- file.path(dirname(ANALYSIS_DIR), "runs", "run_20260412_052731")
STAGE_A <- file.path(RUN_DIR, "stage_a")
STAGE_B <- file.path(RUN_DIR, "stage_b")
STAGE_C <- file.path(RUN_DIR, "stage_c")
DECISIONS <- file.path(RUN_DIR, "decisions")
FIGURES_DIR <- file.path(ANALYSIS_DIR, "figures")
dir.create(FIGURES_DIR, showWarnings = FALSE, recursive = TRUE)

# -- Color scheme --
VARIANT_COLORS <- c(
  baseline = "#4C72B0",
  r1       = "#DD8452",
  r2       = "#55A868",
  r1r2     = "#C44E52"
)
VARIANT_ORDER <- c("baseline", "r1", "r2", "r1r2")
VARIANT_LABELS <- c(
  baseline = "Baseline",
  r1       = "+R1",
  r2       = "+R2",
  r1r2     = "+R1+R2"
)

# -- Theme --
theme_v2 <- theme_minimal(base_size = 12) +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, color = "grey40")
  )
theme_set(theme_v2)

# -- Helper: read YAML decisions --
read_decision <- function(name) {
  yaml::read_yaml(file.path(DECISIONS, paste0(name, ".yaml")))
}

# -- Helper: save figure --
save_fig <- function(p, name, w = 10, h = 7) {
  ggsave(file.path(FIGURES_DIR, paste0(name, ".pdf")), p,
         width = w, height = h, device = cairo_pdf)
  ggsave(file.path(FIGURES_DIR, paste0(name, ".png")), p,
         width = w, height = h, dpi = 300)
}
```

```{r load-data}
# Stage A
df_a1 <- read_csv(file.path(STAGE_A, "a1_coverage_summary.csv"), show_col_types = FALSE)
df_a2 <- read_csv(file.path(STAGE_A, "a2_nhead_summary.csv"), show_col_types = FALSE)
df_a3 <- read_csv(file.path(STAGE_A, "a3_train_params_summary.csv"), show_col_types = FALSE)

# Stage B
df_b <- read_csv(file.path(STAGE_B, "b_screen_summary.csv"), show_col_types = FALSE)

# Stage C
df_c1a <- read_csv(file.path(STAGE_C, "c1a_mode_summary.csv"), show_col_types = FALSE)
df_c1b <- read_csv(file.path(STAGE_C, "c1b_lr_summary.csv"), show_col_types = FALSE)
df_c1c <- read_csv(file.path(STAGE_C, "c1c_head_summary.csv"), show_col_types = FALSE)
df_c2  <- read_csv(file.path(STAGE_C, "c2_final_compare_summary.csv"), show_col_types = FALSE)

# Decisions
shortlist   <- read_decision("shortlist_confirmed")
locked_arch <- read_decision("locked_arch_confirmed")
top3        <- read_decision("stage_a_top3")
reps        <- read_decision("stage_b_representatives")
finals      <- read_decision("final_candidates")

# Factor levels for consistent ordering
for (df_name in c("df_a1", "df_a2", "df_a3", "df_b", "df_c1a", "df_c1b", "df_c1c", "df_c2")) {
  df <- get(df_name)
  if ("model_variant" %in% names(df)) {
    df$model_variant <- factor(df$model_variant, levels = VARIANT_ORDER, labels = VARIANT_LABELS)
    assign(df_name, df)
  }
}
```

# 1. Pipeline Overview {.tabset}

This report presents the complete results of MiCoFormer's V2 hyperparameter search pipeline.
The pipeline follows a three-stage design: **Stage A** (pretrain coarse screening) narrows
112 pretrain trials to 12 candidate checkpoints; **Stage B** (downstream phenotypic prediction)
selects 4 representative checkpoints; **Stage C** (fine-tune parameter search + multi-seed
confirmation) identifies the final two models.

**Locked Architecture:** `r locked_arch$locked_shared_arch_id`
(d_model=`r locked_arch$d_model`, num_layers=`r locked_arch$num_layers`,
nhead=`r locked_arch$nhead`, regime=`r locked_arch$nhead_regime`)

```{r fig12-summary-table}
summary_tbl <- tibble(
  Stage = c("A1: Coverage", "A2: nhead", "A3: Train Params",
            "B: Downstream Screen",
            "C1-A: Pooling/Freeze", "C1-B: Learning Rate", "C1-C: Head Arch",
            "C2: Final Compare"),
  Trials = c(48, 16, 48, 12, 24, 36, 36, 6),
  OK     = c(45, 16, 48, 12, 24, 36, 36, 6),
  Failed = c(3,  0,  0,  0,  0,  0,  0,  0),
  Decision = c(
    "Shortlist 4 architectures",
    "Lock d512_L12_fine",
    "Select top-3 per variant",
    "Select 1 representative per variant",
    "Best pooling_mode + freeze",
    "Best lr_head + lr_encoder",
    "Best head_hidden_dim + head_dropout",
    "Strongest baseline vs final model"
  ),
  Key_Finding = c(
    "d768_L12 OOM in 3/4 variants",
    "Fine regime marginal edge for d512_L12",
    "bs=64, wd=0.1 generally optimal",
    "Pretrain loss != downstream F1",
    "freeze_encoder=True consistently better",
    "lr_encoder irrelevant when frozen",
    "hidden_dim=128 optimal; dropout varies",
    "R2 wins (F1: 0.760 vs 0.698)"
  )
)

summary_tbl %>%
  kbl(caption = "Fig 12: Pipeline Summary",
      col.names = c("Stage", "Trials", "OK", "Failed", "Decision", "Key Finding")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(4, bold = TRUE, background = "#FFF3CD") %>%
  row_spec(8, bold = TRUE, background = "#D4EDDA")
```

---

# 2. Stage A: Pretrain Coarse Screening

## 2.1 Architecture Landscape (Fig 1)

```{r fig01-arch-landscape, fig.width=10, fig.height=7}
# Prepare shortlist for annotation
shortlist_df <- map_dfr(shortlist, as_tibble)

df_a1_plot <- df_a1 %>%
  mutate(
    d_model    = factor(d_model, levels = c(256, 512, 768)),
    num_layers = factor(num_layers, levels = c(4, 6, 8, 12)),
    # Display label: OOM or val_loss value
    cell_label = ifelse(status == "OOM", "OOM", sprintf("%.4f", best_val_loss)),
    # For OOM cells, set val_loss to NA so they get na.value color
    fill_val   = ifelse(status == "OOM", NA_real_, best_val_loss)
  )

# Mark shortlisted architectures
df_a1_plot <- df_a1_plot %>%
  mutate(is_shortlisted = paste(as.numeric(as.character(d_model)),
                                 as.numeric(as.character(num_layers))) %in%
           paste(shortlist_df$d_model, shortlist_df$num_layers))

p1 <- ggplot(df_a1_plot,
             aes(x = num_layers, y = d_model, fill = fill_val)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_tile(data = df_a1_plot %>% filter(is_shortlisted),
            color = "black", linewidth = 1.2, fill = NA) +
  geom_text(aes(label = cell_label), size = 3.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = -1, na.value = "grey80",
                       name = "Val Loss") +
  facet_wrap(~model_variant, ncol = 2) +
  labs(
    title    = "Fig 1: Architecture Landscape (Stage A1 Coverage)",
    subtitle = "Grey = OOM | Black border = shortlisted architectures",
    x = "Number of Layers",
    y = "Model Dimension (d_model)"
  )

print(p1)
save_fig(p1, "fig01_arch_landscape")
```

## 2.2 nhead Regime Comparison (Fig 2)

```{r fig02-nhead-regime, fig.width=9, fig.height=5}
# Combine standard (from a1) and fine (from a2) results for shortlisted architectures
a1_shortlist <- df_a1 %>%
  filter(status == "OK") %>%
  inner_join(shortlist_df %>% select(d_model, num_layers), by = c("d_model", "num_layers")) %>%
  mutate(nhead_regime = "standard")

a2_fine <- df_a2 %>%
  filter(status == "OK") %>%
  mutate(nhead_regime = "fine")

nhead_compare <- bind_rows(
  a1_shortlist %>% select(model_variant, d_model, num_layers, nhead_regime, nhead, best_val_loss),
  a2_fine %>% select(model_variant, d_model, num_layers, nhead_regime, nhead, best_val_loss)
) %>%
  mutate(arch_label = paste0("d", d_model, "_L", num_layers))

# Mean across variants
nhead_mean <- nhead_compare %>%
  group_by(arch_label, nhead_regime) %>%
  summarise(
    mean_val_loss = mean(best_val_loss),
    sd_val_loss   = sd(best_val_loss),
    .groups = "drop"
  ) %>%
  mutate(arch_label = factor(arch_label,
                             levels = c("d512_L12", "d768_L8", "d512_L8", "d768_L6")))

p2 <- ggplot(nhead_mean, aes(x = arch_label, y = mean_val_loss, fill = nhead_regime)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_val_loss - sd_val_loss,
                    ymax = mean_val_loss + sd_val_loss),
                position = position_dodge(0.7), width = 0.2, linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.4f", mean_val_loss)),
            position = position_dodge(0.7), vjust = -0.8, size = 3) +
  scale_fill_manual(values = c(standard = "#7FB3D8", fine = "#E8935A"),
                    name = "nhead Regime",
                    labels = c(standard = "Standard (d_head~64)",
                               fine = "Fine (d_head~32)")) +
  coord_cartesian(ylim = c(2.455, 2.470)) +
  labs(
    title    = "Fig 2: nhead Regime Comparison (Stage A2)",
    subtitle = "Mean val_loss across 4 model variants | Error bars = SD",
    x = "Architecture",
    y = "Validation Loss"
  )

print(p2)
save_fig(p2, "fig02_nhead_regime", w = 9, h = 5)
```

## 2.3 Training Parameter Grid (Fig 3)

```{r fig03-train-param-grid, fig.width=10, fig.height=7}
df_a3_plot <- df_a3 %>%
  filter(status == "OK") %>%
  mutate(
    lr_label = sprintf("%.0e", lr),
    bs_wd    = paste0("bs=", batch_size, "\nwd=", weight_decay),
    bs_wd    = factor(bs_wd, levels = c("bs=64\nwd=0.05", "bs=64\nwd=0.1",
                                         "bs=128\nwd=0.05", "bs=128\nwd=0.1"))
  )

p3 <- ggplot(df_a3_plot, aes(x = bs_wd, y = lr_label, fill = best_val_loss)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.4f", best_val_loss)), size = 2.8) +
  scale_fill_distiller(palette = "YlOrRd", direction = -1, name = "Val Loss") +
  facet_wrap(~model_variant, ncol = 2) +
  labs(
    title    = "Fig 3: Training Parameter Grid (Stage A3)",
    subtitle = "Locked architecture: d512_L12_fine | 3 x 2 x 2 = 12 grid points per variant",
    x = "Batch Size / Weight Decay",
    y = "Learning Rate"
  )

print(p3)
save_fig(p3, "fig03_train_param_grid")
```

---

# 3. Stage B: Downstream Screening

## 3.1 Pretrain-Downstream Disconnect (Fig 4)

```{r fig04-disconnect, fig.width=12, fig.height=5}
# Left panel: pretrain val_loss (best rank1 per variant from stage_a_top3)
pretrain_best <- map_dfr(top3, as_tibble) %>%
  filter(selected_rank == 1) %>%
  select(model_variant, best_val_loss) %>%
  mutate(model_variant = factor(model_variant, levels = VARIANT_ORDER, labels = VARIANT_LABELS))

# Right panel: downstream F1 (best per variant from stage_b_representatives)
downstream_best <- map_dfr(reps, as_tibble) %>%
  select(model_variant, val_macro_f1) %>%
  mutate(model_variant = factor(model_variant, levels = VARIANT_ORDER, labels = VARIANT_LABELS))

# Left: pretrain loss (lower is "better", so we order accordingly)
p4_left <- ggplot(pretrain_best,
                  aes(x = reorder(model_variant, best_val_loss),
                      y = best_val_loss, fill = model_variant)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.4f", best_val_loss)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = VARIANT_COLORS[VARIANT_ORDER],
                    breaks = VARIANT_LABELS) +
  coord_cartesian(ylim = c(2.44, 2.46)) +
  labs(
    title = "Pretrain Val Loss (lower = better)",
    subtitle = "Stage A top-1 checkpoint",
    x = NULL, y = "Val Loss"
  )

# Right: downstream F1 (higher is better)
p4_right <- ggplot(downstream_best,
                   aes(x = reorder(model_variant, -val_macro_f1),
                       y = val_macro_f1, fill = model_variant)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = VARIANT_COLORS[VARIANT_ORDER],
                    breaks = VARIANT_LABELS) +
  coord_cartesian(ylim = c(0, 0.8)) +
  labs(
    title = "Downstream Macro F1 (higher = better)",
    subtitle = "Stage B best representative",
    x = NULL, y = "Macro F1"
  )

p4 <- p4_left + p4_right +
  plot_annotation(
    title    = "Fig 4: The Pretrain-Downstream Disconnect",
    subtitle = "Best pretrain loss (left) does NOT predict best downstream F1 (right) — rank inversion between +R1 and +R2",
    theme = theme(plot.title = element_text(face = "bold", size = 14),
                  plot.subtitle = element_text(size = 11, color = "grey30"))
  )

print(p4)
save_fig(p4, "fig04_pretrain_downstream_disconnect", w = 12, h = 5)
```

## 3.2 Pretrain Rank vs Downstream F1 (Fig 5)

```{r fig05-rank-vs-f1, fig.width=8, fig.height=5}
df_b_plot <- df_b %>%
  filter(status == "OK") %>%
  mutate(
    pretrain_rank = as.integer(str_extract(trial_id, "\\d+$")),
    model_variant = factor(model_variant, levels = VARIANT_LABELS)
  )

p5 <- ggplot(df_b_plot, aes(x = pretrain_rank, y = val_macro_f1,
                             color = model_variant, group = model_variant)) +
  geom_line(linewidth = 0.8, alpha = 0.5) +
  geom_point(size = 3.5) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)),
            vjust = -1, size = 3, show.legend = FALSE) +
  scale_color_manual(values = VARIANT_COLORS[VARIANT_ORDER],
                     breaks = VARIANT_LABELS, name = "Variant") +
  scale_x_continuous(breaks = 1:3, labels = c("Rank 1", "Rank 2", "Rank 3")) +
  labs(
    title    = "Fig 5: All 12 Stage B Trials",
    subtitle = "Pretrain checkpoint rank does not reliably predict downstream performance",
    x = "Pretrain Checkpoint Rank (by val_loss)",
    y = "Downstream Macro F1"
  )

print(p5)
save_fig(p5, "fig05_rank_vs_f1", w = 8, h = 5)
```

---

# 4. Stage C: Fine-tune Parameter Search

## 4.1 Pooling Mode x Freeze Encoder (Fig 6)

```{r fig06-pooling-freeze, fig.width=11, fig.height=7}
df_c1a_plot <- df_c1a %>%
  filter(status == "OK") %>%
  mutate(
    freeze_label = ifelse(freeze_encoder, "Frozen", "Unfrozen"),
    freeze_label = factor(freeze_label, levels = c("Frozen", "Unfrozen")),
    pooling_mode = factor(pooling_mode,
                          levels = c("sample", "mean_pool", "sample_and_mean"),
                          labels = c("Sample", "Mean Pool", "Sample+Mean"))
  )

p6 <- ggplot(df_c1a_plot, aes(x = pooling_mode, y = val_macro_f1,
                               fill = freeze_label)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)),
            position = position_dodge(0.7), vjust = -0.5, size = 2.8) +
  scale_fill_manual(values = c(Frozen = "#55A868", Unfrozen = "#DD8452"),
                    name = "Encoder") +
  facet_wrap(~model_variant, ncol = 2, scales = "free_y") +
  labs(
    title    = "Fig 6: Pooling Mode x Freeze Encoder (Stage C1-A)",
    subtitle = "Frozen encoder consistently outperforms unfrozen across all variants",
    x = "Pooling Mode",
    y = "Macro F1"
  )

print(p6)
save_fig(p6, "fig06_pooling_freeze", w = 11, h = 7)
```

## 4.2 Learning Rate Grid (Fig 7)

```{r fig07-lr-grid, fig.width=10, fig.height=7}
df_c1b_plot <- df_c1b %>%
  filter(status == "OK") %>%
  mutate(
    lr_head_label    = sprintf("%.0e", lr_head),
    lr_encoder_label = sprintf("%.0e", lr_encoder)
  )

p7 <- ggplot(df_c1b_plot,
             aes(x = lr_encoder_label, y = lr_head_label, fill = val_macro_f1)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)), size = 3) +
  scale_fill_distiller(palette = "Greens", direction = 1, name = "Macro F1") +
  facet_wrap(~model_variant, ncol = 2) +
  labs(
    title    = "Fig 7: Learning Rate Grid (Stage C1-B)",
    subtitle = "Uniform columns confirm lr_encoder is irrelevant when encoder is frozen",
    x = "Encoder Learning Rate",
    y = "Head Learning Rate"
  )

print(p7)
save_fig(p7, "fig07_lr_grid")
```

## 4.3 Head Architecture Grid (Fig 8)

```{r fig08-head-grid, fig.width=10, fig.height=7}
df_c1c_plot <- df_c1c %>%
  filter(status == "OK") %>%
  mutate(
    hd_label = factor(head_hidden_dim, levels = c(0, 128, 256),
                      labels = c("0 (linear)", "128", "256")),
    dp_label = factor(head_dropout)
  )

p8 <- ggplot(df_c1c_plot,
             aes(x = dp_label, y = hd_label, fill = val_macro_f1)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)), size = 3) +
  scale_fill_distiller(palette = "Greens", direction = 1, name = "Macro F1") +
  facet_wrap(~model_variant, ncol = 2) +
  labs(
    title    = "Fig 8: Head Architecture Grid (Stage C1-C)",
    subtitle = "hidden_dim=128 is consistently optimal; dropout preference varies by variant",
    x = "Head Dropout",
    y = "Head Hidden Dimension"
  )

print(p8)
save_fig(p8, "fig08_head_grid")
```

## 4.4 Progressive Improvement (Fig 9)

```{r fig09-waterfall, fig.width=9, fig.height=5}
# Collect best F1 per variant at each stage
best_b <- df_b %>%
  filter(status == "OK") %>%
  group_by(model_variant) %>%
  slice_max(val_macro_f1, n = 1) %>%
  ungroup() %>%
  select(model_variant, val_macro_f1) %>%
  mutate(stage = "B: Screen")

best_c1a <- df_c1a %>%
  filter(status == "OK") %>%
  group_by(model_variant) %>%
  slice_max(val_macro_f1, n = 1) %>%
  ungroup() %>%
  select(model_variant, val_macro_f1) %>%
  mutate(stage = "C1-A: Mode")

best_c1b <- df_c1b %>%
  filter(status == "OK") %>%
  group_by(model_variant) %>%
  slice_max(val_macro_f1, n = 1) %>%
  ungroup() %>%
  select(model_variant, val_macro_f1) %>%
  mutate(stage = "C1-B: LR")

best_c1c <- df_c1c %>%
  filter(status == "OK") %>%
  group_by(model_variant) %>%
  slice_max(val_macro_f1, n = 1) %>%
  ungroup() %>%
  select(model_variant, val_macro_f1) %>%
  mutate(stage = "C1-C: Head")

waterfall <- bind_rows(best_b, best_c1a, best_c1b, best_c1c) %>%
  mutate(
    stage = factor(stage, levels = c("B: Screen", "C1-A: Mode",
                                      "C1-B: LR", "C1-C: Head"))
  )

p9 <- ggplot(waterfall, aes(x = stage, y = val_macro_f1,
                             color = model_variant, group = model_variant)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.3f", val_macro_f1)),
            vjust = -1, size = 3, show.legend = FALSE) +
  scale_color_manual(values = VARIANT_COLORS[VARIANT_ORDER],
                     breaks = VARIANT_LABELS, name = "Variant") +
  labs(
    title    = "Fig 9: Progressive Improvement Through Fine-tune Search",
    subtitle = "Biggest jump from B to C1-A (switching to frozen encoder)",
    x = "Search Stage",
    y = "Best Macro F1"
  )

print(p9)
save_fig(p9, "fig09_progressive_improvement", w = 9, h = 5)
```

---

# 5. Final Results & Caveats

## 5.1 Multi-Seed Final Comparison (Fig 10)

```{r fig10-multi-seed, fig.width=10, fig.height=5}
df_c2_plot <- df_c2 %>%
  filter(status == "OK") %>%
  mutate(
    role_label = ifelse(role == "strongest_baseline",
                        "Strongest Baseline", "Strongest Final (+R2)"),
    role_label = factor(role_label, levels = c("Strongest Baseline",
                                                "Strongest Final (+R2)"))
  )

# Summary stats
c2_summary <- df_c2_plot %>%
  group_by(role_label) %>%
  summarise(
    mean_f1   = mean(val_macro_f1),
    sd_f1     = sd(val_macro_f1),
    mean_auroc = mean(val_auroc),
    sd_auroc   = sd(val_auroc),
    .groups = "drop"
  )

# F1 panel
p10_f1 <- ggplot() +
  geom_col(data = c2_summary,
           aes(x = role_label, y = mean_f1, fill = role_label),
           width = 0.5, alpha = 0.7) +
  geom_errorbar(data = c2_summary,
                aes(x = role_label, ymin = mean_f1 - sd_f1, ymax = mean_f1 + sd_f1),
                width = 0.15, linewidth = 0.6) +
  geom_jitter(data = df_c2_plot,
              aes(x = role_label, y = val_macro_f1),
              width = 0.05, size = 2.5, alpha = 0.8) +
  geom_text(data = c2_summary,
            aes(x = role_label,
                y = mean_f1 + sd_f1 + 0.02,
                label = sprintf("%.3f +/- %.3f", mean_f1, sd_f1)),
            size = 3.5) +
  scale_fill_manual(values = c("Strongest Baseline" = "#4C72B0",
                                "Strongest Final (+R2)" = "#55A868")) +
  coord_cartesian(ylim = c(0.5, 0.95)) +
  labs(title = "Macro F1", x = NULL, y = "Macro F1") +
  theme(legend.position = "none")

# AUROC panel
p10_auroc <- ggplot() +
  geom_col(data = c2_summary,
           aes(x = role_label, y = mean_auroc, fill = role_label),
           width = 0.5, alpha = 0.7) +
  geom_errorbar(data = c2_summary,
                aes(x = role_label, ymin = mean_auroc - sd_auroc,
                    ymax = mean_auroc + sd_auroc),
                width = 0.15, linewidth = 0.6) +
  geom_jitter(data = df_c2_plot,
              aes(x = role_label, y = val_auroc),
              width = 0.05, size = 2.5, alpha = 0.8) +
  geom_text(data = c2_summary,
            aes(x = role_label,
                y = mean_auroc + sd_auroc + 0.01,
                label = sprintf("%.3f +/- %.3f", mean_auroc, sd_auroc)),
            size = 3.5) +
  scale_fill_manual(values = c("Strongest Baseline" = "#4C72B0",
                                "Strongest Final (+R2)" = "#55A868")) +
  coord_cartesian(ylim = c(0.7, 1.0)) +
  labs(title = "AUROC", x = NULL, y = "AUROC") +
  theme(legend.position = "none")

p10 <- p10_f1 + p10_auroc +
  plot_annotation(
    title    = "Fig 10: Multi-Seed Final Comparison (Stage C2)",
    subtitle = "3 seeds (42, 52, 62) | Bars = mean, error bars = SD, dots = individual seeds",
    theme = theme(plot.title = element_text(face = "bold", size = 14),
                  plot.subtitle = element_text(size = 11, color = "grey30"))
  )

print(p10)
save_fig(p10, "fig10_multi_seed_comparison", w = 10, h = 5)
```

## 5.2 Validation vs Test Gap (Fig 11)

```{r fig11-val-test-gap, fig.width=9, fig.height=5}
df_gap <- df_c2_plot %>%
  select(trial_id, role_label, seed, val_macro_f1, test_macro_f1) %>%
  mutate(
    trial_label = paste0(role_label, " (seed ", seed, ")"),
    trial_label = factor(trial_label,
                         levels = rev(unique(trial_label)))
  )

df_gap_long <- df_gap %>%
  pivot_longer(cols = c(val_macro_f1, test_macro_f1),
               names_to = "split", values_to = "f1") %>%
  mutate(split = ifelse(split == "val_macro_f1", "Validation", "Test"))

p11 <- ggplot(df_gap, aes(y = trial_label)) +
  # Connecting segments
  geom_segment(aes(x = test_macro_f1, xend = val_macro_f1,
                   yend = trial_label, color = role_label),
               linewidth = 1.2, alpha = 0.5) +
  # Points
  geom_point(data = df_gap_long,
             aes(x = f1, y = trial_label, shape = split),
             size = 3.5) +
  # Labels for val
  geom_text(aes(x = val_macro_f1, label = sprintf("%.3f", val_macro_f1)),
            hjust = -0.3, size = 3) +
  # Labels for test
  geom_text(aes(x = test_macro_f1, label = sprintf("%.3f", test_macro_f1)),
            hjust = 1.3, size = 3) +
  scale_color_manual(values = c("Strongest Baseline" = "#4C72B0",
                                 "Strongest Final (+R2)" = "#55A868"),
                     guide = "none") +
  scale_shape_manual(values = c(Validation = 16, Test = 17), name = "Split") +
  labs(
    title    = "Fig 11: Validation vs Test Gap",
    subtitle = "Consistent and large F1 drop from validation to test across all seeds and models",
    x = "Macro F1",
    y = NULL
  ) +
  theme(legend.position = "right")

print(p11)
save_fig(p11, "fig11_val_test_gap", w = 9, h = 5)
```

---

# 6. Conclusions

**Key takeaways from the V2 pipeline:**

1. **Architecture**: d512_L12 with fine nhead regime (16 heads) was selected as the locked shared architecture.

2. **The central finding**: Pretrain val_loss does NOT predict downstream classification performance.
   +R1 achieved the best pretrain loss but the worst downstream F1. +R2 achieved mediocre pretrain
   loss but the best downstream F1. This validates the multi-stage design.

3. **R2 wins**: The taxonomy distance attention bias (+R2) is the most impactful module for
   downstream phenotype prediction (mean F1: 0.760 vs baseline 0.698).

4. **Frozen encoder**: Freezing the pretrained encoder consistently outperforms fine-tuning it,
   suggesting the pretrained representations are already high-quality for this task.

5. **Caveat — Val/Test gap**: All models show a substantial F1 drop from validation to test
   (0.17--0.33), indicating potential distribution shift or overfitting. This requires further
   investigation before drawing final conclusions.

6. **Seed variance**: R2 shows higher seed variance (std=0.064 vs baseline 0.047), suggesting
   the advantage is real but not as stable as desired.
